### 2024-05-14 thesis/input_files/probabilistic.tex
May 13th 2024

% In our implementatiton we take the point that minsimizes the distance between the boundary of domain of alpha, in this way we are guaranteed we do not take ith index of alpha_i for retrieving b, because it that case it would not work.
%contribution is implementation kernel quantile regression for pythos users since there exists only r version
\subsection{Weather quantiles}
In order to get acquainted with the inner workings of the presented methods, this section covers an application explaining practical details and comparing results.
\\
The dataset used is the Melbourne daily maximum temperatures \cite{hdrcde}. It contains the daily maximum temperatures in Melbourne, Australia, from 1981-1990, excluding leap days, see figure \ref{fig:melbourne_temperature_data}.
Due to the bimodality of the data, such dataset is commonly used to give a difficult quantile regression problem \cite{hyndman1996estimating}, thus why we chose it. The observed bimodality is that a hot day is likely to be followed by either an equally hot day or one much cooler.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/melbourne_temperature.png}
    \caption{Melbourne temperatures dataset}
    \label{fig:melbourne_temperature_data}
\end{figure}
Hereafter, the results of the four presented methods on the Melbourne dataset are reported, see figure \ref{fig:melbourne_quantiles_comparison} for a visualisation. Hyperparameter tuning have been optimised through cross validation, see \ref{appendix:cross_validation}.
\begin{figure}[!h]
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_linear_quantile_regression.png} 
      \caption{Linear quantile regression} 
      \label{fig:melbourne_linear_quantile_regression} 
      \vspace{4ex}
    \end{subfigure}%% 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_gradient_boosting_quantile_regression.png} 
      \caption{Gradient boosting quantile regression} 
      \label{fig:melbourne_gradient_boosting_quantile_regression} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_quantile_forest.png} 
      \caption{Quantile forest} 
      \label{fig:melbourne_quantile_forest} 
    \end{subfigure}%%
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_kernel_quantile_regression.png}
      \caption{Kernel quantile regression} 
      \label{fig:melborune_kernel_quantile_regression} 
    \end{subfigure} 
    \caption{Quantile regressors}
    \label{fig:melbourne_quantiles_comparison} 
  \end{figure}

@@

\begin{table}[!h]
    \caption{Pinball loss quantile-wise Melbourne data}
    \begin{tabular}{lllll}
    \toprule
    Quantiles & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    0.100000 & 0.710644 & 0.549232 & 0.562888 & 0.540235 \\
    0.200000 & 1.155014 & 0.938561 & 0.946712 & 0.903213 \\
    0.300000 & 1.417805 & 1.212671 & 1.222407 & 1.173803 \\
    0.400000 & 1.540108 & 1.399925 & 1.409293 & 1.367943 \\
    0.500000 & 1.574957 & 1.517281 & 1.484589 & 1.455849 \\
    0.600000 & 1.525114 & 1.498608 & 1.495474 & 1.447470 \\
    0.700000 & 1.397918 & 1.372183 & 1.362173 & 1.331503 \\
    0.800000 & 1.170140 & 1.115077 & 1.123195 & 1.096606 \\
    0.900000 & 0.787195 & 0.714075 & 0.734112 & 0.715085 \\
    \bottomrule
    \end{tabular}
\end{table}
As already pointed out, the quantile regression with q=0.5 corresponds to the standard regression problem, hence we can compare the proposed methods also in terms of the mean absolute error.
\begin{table}[!h]
\caption{Mean absolute error Melbourne data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    MAE & 3.253882 & 3.134805 & 3.095041 & 3.024336 \\
    \bottomrule
    \end{tabular}
\end{table}  
From these tables, we can see that kernel quantile regression outperformed the simple quantile regressor as well as the more complex models like quantile forest and gradient boosting quantile regression for the Melbourne temperatures dataset. Not only kernel quantile regression was the best in terms of total pinball loss but also the best in terms of each quantile pinball loss and mean absolute error.
Comparison has been carried out on further datasets, yielding similar conclusion as the one of above, to know more see \ref{appendix:quantile_regressor_extensive_comparison}.

@@

% tables
\begin{table}[!h]
    \caption{Kernels comparison pinball loss Melbourne data}
    \label{tab:kernel pinball comparison}
    \begin{center}
    \begin{tabular}{lll}
        \toprule
        & Kernel & Pinball loss
        \\
        \midrule
        & Gauss RBF &  10.031708 \\
        & Laplacian &  10.056884 \\
        & Gauss RBFxLaplacian & 10.150826 \\
        & Cosine & 16.253973    \\
        & Linear & 10.463867    \\
        & Polynomial & 11.238393     \\
        & Sigmoid & 16.253973            \\
        & Chi squared & 10.023732       \\
        & Matern & \textbf{10.021369}  \\
        & Periodic  & 15.946272\\
        \bottomrule
        \end{tabular}
    \end{center}
    \end{table}

\begin{table}[!h]
    \caption{Kernels comparison pinball loss quantile-wise Melbourne data}
    \label{tab:kernel pinball comparison quantile-wise}
    \begin{adjustbox}{width=\textwidth}
    \begin{tabular}{lllllllllll}
    \toprule
    Kernel/Quantiles & 0.1 & 0.2 & 0.3 & 0.4 & 0.5 & 0.6 & 0.7 & 0.8 & 0.9 \\
    \midrule
    Gauss RBF &  \textbf{0.540145} &
    \textbf{0.903193} &
    1.173627 &
    1.368022 &
    1.456053 &
    1.447470 &
    \textbf{1.331375} &
    1.096649 &
    0.715013 \\
    Laplacian &  0.542499 &
    0.906517 &
    1.177309 &
    1.371198 &
    1.462111 &
    1.453119 &
    1.340193 &
    \textbf{1.095991} &
    0.707947 \\
    Gauss RBFxLaplacian &0.545787 &
    0.915559 &
    1.189965 &
    1.387109 &
    1.479901 &
    1.475541 &
    1.350803 &
    1.101151 &
    \textbf{0.704922} \\
    Cosine & 0.755479 &
    1.343973 &
    1.802603 &
    2.123425 &
    2.307123 &
    2.367123 &
    2.262603 &
    1.971644 &
    1.320411 \\
    Linear &  0.565663 &
    0.947889 &
    1.250009 &
    1.437863 &
    1.506723 &
    1.478379 &
    1.358865 &
    1.135702 &
    0.782660   \\
    Polynomial &   0.542573 &
    0.908681 &
    1.178357 &
    \textbf{1.362351} &
    1.450249 &
    1.455228 &
    1.339822 &
    1.107768 &
    0.744118  \\
    Sigmoid &   0.755479 &
    1.343973 &
    1.802603 &
    2.123425 &
    2.307123 &
    2.367123 &
    2.262603 &
    1.971644 &
    1.320411          \\
    Chi squared &   0.541430 &
    0.906352 &
    1.173164 &
    1.358862 &
    \textbf{1.444130} &
    \textbf{1.440915} &
    1.332140 &
    1.100262 &
    0.726625     \\
    Matern &  0.541192 &
    0.903209 &
    \textbf{1.169543} &
    1.364430 &
    1.451248 &
    1.442511 &
    1.332522 &
    1.096750 &
    0.716226 \\
    Periodic  & 0.759452 &
    1.352747 &
    1.809921 &
    2.125765 &
    2.314718 &
    2.379657 &
    2.274711 &
    1.976768 &
    1.328630 \\
    \bottomrule
    \end{tabular}
    \end{adjustbox}
\end{table}

\begin{table}[!h]
    \caption{Kernels comparison mean absolute error Melbourne data}
    \label{tab:kernel mae comparison}
    \begin{center}
    \begin{tabular}{lll}
        \toprule
         & Kernel & MAE \\
        \midrule
        & Gauss RBF &  3.024336 \\
        & Laplacian &  3.038633 \\
        & Gauss RBFxLaplacian &  3.070812\\
        & Cosine &    4.849589 \\
        & Linear &     3.142352\\
        & Polynomial &     3.064472 \\
        & Sigmoid &      4.849589       \\
        & Chi squared &    \textbf{3.022778}    \\
        & Matern &  3.024517 \\
        & Periodic  & 4.880673 \\
        \bottomrule
        \end{tabular}
    \end{center}
    \end{table}

@@

\begin{figure}[!h]\ContinuedFloat
    \begin{subfigure}[b]{0.5\linewidth}
        \centering
        \includegraphics[width=1.1\textwidth]{images/melborune_sigmoid_kernel_quantile_regression.png}
        \caption{Sigmoid} 
        \label{} 
        \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
        \centering
        \includegraphics[width=1.1\textwidth]{images/melborune_chi_squared_kernel_quantile_regression.png}
        \caption{Chi squared} 
        \label{} 
        \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
        \centering
        \includegraphics[width=1.1\textwidth]{images/melborune_matern_kernel_quantile_regression.png}
        \caption{Matern} 
        \label{} 
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
        \centering
        \includegraphics[width=1.1\textwidth]{images/melborune_periodic_kernel_quantile_regression.png}
        \caption{Periodic} 
        \label{} 
    \end{subfigure} 
    \caption{Kernel quantile regressors comparison}
    \label{fig:kernel quantile regressors comparison} 
\end{figure}

>>>>

% In our implementatiton we take the point that minsimizes the distance between the boundary of domain of alpha, in this way we are guaranteed we do not take ith index of alpha_i for retrieving b, because it that case it would not work.
%contribution is implementation kernel quantile regression for pythos users since there exists only r version
\subsection{Weather quantiles}
In order to get acquainted with the inner workings of the presented methods, this section covers an application explaining practical details and comparing results.
\\
The dataset used is the Melbourne daily maximum temperatures \cite{hdrcde}. It contains the daily maximum temperatures in Melbourne, Australia, from 1981-1990, excluding leap days, see figure \ref{fig:melbourne_temperature_data}.
Due to the bimodality of the data, such dataset is commonly used to give a difficult quantile regression problem \cite{hyndman1996estimating}, thus why we chose it. The observed bimodality is that a hot day is likely to be followed by either an equally hot day or one much cooler.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/melbourne_temperature.png}
    \caption{Melbourne temperatures dataset}
    \label{fig:melbourne_temperature_data}
\end{figure}
Hereafter, the results of the four presented methods on the Melbourne dataset are reported, see figure \ref{fig:melbourne_quantiles_comparison} for a visualisation. Notice, the kernel considered here is the gaussian rbf. Hyperparameter tuning have been optimised through cross validation, see \ref{appendix:cross_validation}.
\begin{figure}[!h]
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_linear_quantile_regression.png} 
      \caption{Linear quantile regression} 
      \label{fig:melbourne_linear_quantile_regression} 
      \vspace{4ex}
    \end{subfigure}%% 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_gradient_boosting_quantile_regression.png} 
      \caption{Gradient boosting quantile regression} 
      \label{fig:melbourne_gradient_boosting_quantile_regression} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_quantile_forest.png} 
      \caption{Quantile forest} 
      \label{fig:melbourne_quantile_forest} 
    \end{subfigure}%%
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_kernel_quantile_regression.png}
      \caption{Kernel quantile regression} 
      \label{fig:melborune_kernel_quantile_regression} 
    \end{subfigure} 
    \caption{Quantile regressors}
    \label{fig:melbourne_quantiles_comparison} 
  \end{figure}

@@

\begin{table}[!h]
    \caption{Pinball loss quantile-wise Melbourne data}
    \begin{tabular}{lllll}
    \toprule
    Quantiles & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    0.100000 & 0.710644 & 0.549232 & 0.562888 & \textbf{0.540235} \\
    0.200000 & 1.155014 & 0.938561 & 0.946712 & \textbf{0.903213} \\
    0.300000 & 1.417805 & 1.212671 & 1.222407 & \textbf{1.173803} \\
    0.400000 & 1.540108 & 1.399925 & 1.409293 & \textbf{1.367943} \\
    0.500000 & 1.574957 & 1.517281 & 1.484589 & \textbf{1.455849} \\
    0.600000 & 1.525114 & 1.498608 & 1.495474 & 1.\textbf{447470} \\
    0.700000 & 1.397918 & 1.372183 & 1.362173 & \textbf{1.331503} \\
    0.800000 & 1.170140 & 1.115077 & 1.123195 & \textbf{1.096606} \\
    0.900000 & 0.787195 & \textbf{0.714075} & 0.734112 & 0.715085 \\
    \bottomrule
    \end{tabular}
\end{table}
As already pointed out, the quantile regression with q=0.5 corresponds to the standard regression problem, hence we can compare the proposed methods also in terms of the mean absolute error.
\begin{table}[!h]
\caption{Mean absolute error Melbourne data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    MAE & 3.253882 & 3.134805 & 3.095041 & \textbf{3.024336} \\
    \bottomrule
    \end{tabular}
\end{table}  
From these tables, we can see that kernel quantile regression outperformed the simple quantile regressor as well as the more complex models like quantile forest and gradient boosting quantile regression for the Melbourne temperatures dataset. Not only kernel quantile regression was the best in terms of total pinball loss but also the best in terms of each quantile pinball loss and mean absolute error.
Comparison has been carried out on further datasets, yielding similar conclusion as the one of above, to know more see \ref{appendix:quantile_regressor_extensive_comparison}.

@@

% tables
\begin{table}[!h]
    \caption{Kernels comparison pinball loss Melbourne data}
    \label{tab:kernel pinball comparison}
    \begin{center}
    \begin{tabular}{lll}
        \toprule
        & Kernel & Pinball loss
        \\
        \midrule
        & Gauss RBF &  10.031708 \\
        & Laplacian &  10.056884 \\
        & Gauss RBFxLaplacian & 10.150826 \\
        & Cosine & 16.253973    \\
        & Linear & 10.463867    \\
        & Polynomial & 11.238393     \\
        & Sigmoid & 16.253973            \\
        & Chi squared & \textbf{10.023732}       \\
        & Matern 0.5 & 10.056883  \\
        & Matern 1.5 & 10.040288  \\
        & Matern 2.5 & 10.024495    \\
        & Periodic  & 15.946272\\
        \bottomrule
        \end{tabular}
    \end{center}
    \end{table}

\begin{table}[!h]
    \caption{Kernels comparison pinball loss quantile-wise Melbourne data}
    \label{tab:kernel pinball comparison quantile-wise}
    \begin{adjustbox}{width=\textwidth}
    \begin{tabular}{lllllllllll}
    \toprule
    Kernel/Quantiles & 0.1 & 0.2 & 0.3 & 0.4 & 0.5 & 0.6 & 0.7 & 0.8 & 0.9 \\
    \midrule
    Gauss RBF &  \textbf{0.540145} &
    0.903193 &
    1.173627 &
    1.368022 &
    1.456053 &
    1.447470 &
    1.331375 &
    1.096649 &
    0.715013 \\
    Laplacian &  0.542499 &
    0.906517 &
    1.177309 &
    1.371198 &
    1.462111 &
    1.453119 &
    1.340193 &
    1.095991 &
    0.707947 \\
    Gauss RBFxLaplacian &0.545787 &
    0.915559 &
    1.189965 &
    1.387109 &
    1.479901 &
    1.475541 &
    1.350803 &
    1.101151 &
    \textbf{0.704922} \\
    Cosine & 0.755479 &
    1.343973 &
    1.802603 &
    2.123425 &
    2.307123 &
    2.367123 &
    2.262603 &
    1.971644 &
    1.320411 \\
    Linear &  0.565663 &
    0.947889 &
    1.250009 &
    1.437863 &
    1.506723 &
    1.478379 &
    1.358865 &
    1.135702 &
    0.782660   \\
    Polynomial &   0.542573 &
    0.908681 &
    1.178357 &
    1.362351 &
    1.450249 &
    1.455228 &
    1.339822 &
    1.107768 &
    0.744118  \\
    Sigmoid &   0.755479 &
    1.343973 &
    1.802603 &
    2.123425 &
    2.307123 &
    2.367123 &
    2.262603 &
    1.971644 &
    1.320411          \\
    Chi squared &   0.541430 &
    0.906352 &
    1.173164 &
    \textbf{1.358862} &
    \textbf{1.444130} &
    \textbf{1.440915} &
    1.332140 &
    1.100262 &
    0.726625     \\
    Matern 0.5 &  0.542499 &
    0.906517 &
    1.177309 &
    1.371198 &
    1.462111 &
    1.453119 &
    1.340193 &
    1.095991 &
    0.707947 \\
    Matern 1.5 &  0.541410 &
    0.902901 &
    1.173943 &
    1.371218 &
    1.462212 &
    1.449296 &
    1.333655 &
    \textbf{1.095092} &
    0.710560 \\
    Matern 2.5 &  0.541233 &
    \textbf{0.902231} &
    \textbf{1.172001} &
    1.366449 &
    1.456505 &
    1.444124 &
    \textbf{1.331251} &
    1.095674 &
    0.715029 \\
    Periodic  & 0.759452 &
    1.352747 &
    1.809921 &
    2.125765 &
    2.314718 &
    2.379657 &
    2.274711 &
    1.976768 &
    1.328630 \\
    \bottomrule
    \end{tabular}
    \end{adjustbox}
\end{table}

\begin{table}[!h]
    \caption{Kernels comparison mean absolute error Melbourne data}
    \label{tab:kernel mae comparison}
    \begin{center}
    \begin{tabular}{lll}
        \toprule
         & Kernel & MAE \\
        \midrule
        & Gauss RBF &  3.024336 \\
        & Laplacian &  3.038633 \\
        & Gauss RBFxLaplacian &  3.070812\\
        & Cosine &    4.849589 \\
        & Linear &     3.142352\\
        & Polynomial &     3.064472 \\
        & Sigmoid &      4.849589       \\
        & Chi squared &    3.022778    \\
        & Matern 0.5 &  3.038633 \\
        & Matern 1.5 &  3.023060 \\
        & Matern 2.5 &  \textbf{3.021949} \\
        & Periodic  & 4.880673 \\
        \bottomrule
        \end{tabular}
    \end{center}
    \end{table}

@@

\begin{figure}[!h]\ContinuedFloat
    \begin{subfigure}[b]{0.5\linewidth}
        \centering
        \includegraphics[width=1.1\textwidth]{images/melborune_sigmoid_kernel_quantile_regression.png}
        \caption{Sigmoid} 
        \label{} 
        \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
        \centering
        \includegraphics[width=1.1\textwidth]{images/melborune_chi_squared_kernel_quantile_regression.png}
        \caption{Chi squared} 
        \label{} 
        \vspace{4ex}
    \end{subfigure}

    \begin{subfigure}[b]{0.5\linewidth}
        \centering
        \includegraphics[width=1.1\textwidth]{images/melborune_matern_0.5_kernel_quantile_regression.png}
        \caption{Matern 0.5} 
        \label{} 
        \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
        \centering
        \includegraphics[width=1.1\textwidth]{images/melborune_matern_1.5_kernel_quantile_regression.png}
        \caption{Matern 1.5} 
        \label{} 
        \vspace{4ex}
    \end{subfigure}

    \begin{subfigure}[b]{0.5\linewidth}
        \centering
        \includegraphics[width=1.1\textwidth]{images/melborune_matern_2.5_kernel_quantile_regression.png}
        \caption{Matern 2.5} 
        \label{} 
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
        \centering
        \includegraphics[width=1.1\textwidth]{images/melborune_periodic_kernel_quantile_regression.png}
        \caption{Periodic} 
        \label{} 
    \end{subfigure} 
    \caption{Kernel quantile regressors comparison}
    \label{fig:kernel quantile regressors comparison} 
\end{figure}

### 2024-05-14 thesis/input_files/symbols.tex
April 6th 2024
May 13th 2024

%E
\glsxtrnewsymbol[description={Exploratory data analysis}]{EDA}{EDA}
\glsxtrnewsymbol[description={Electricity forecasting}]{EF}{EF}
\glsxtrnewsymbol[description={Electricity load forecasting}]{ELF}{ELF}
\glsxtrnewsymbol[description={Empirical mode decomposition}]{EMD}{EMD}
\glsxtrnewsymbol[description={Electricity load forecasting}]{EPF}{EPF}
\glsxtrnewsymbol[description={Exponential smoothing models}]{ESM}{ESM}
\glsxtrnewsymbol[description={Extract transform load}]{ETL}{ETL}
\glsxtrnewsymbol[description={Electric vehicles}]{EVs}{EVs}

@@

>>>>

%E
\glsxtrnewsymbol
[description={European Centre for Medium-Range Weather Forecasts}]{ECMWF}{ECMWF}
[description={Exploratory data analysis}]{EDA}{EDA}
\glsxtrnewsymbol[description={Electricity forecasting}]{EF}{EF}
\glsxtrnewsymbol[description={Electricity load forecasting}]{ELF}{ELF}
\glsxtrnewsymbol[description={Empirical mode decomposition}]{EMD}{EMD}
\glsxtrnewsymbol[description={European network of transmission system operators for electricity}]{ENTSOE}{ENTSOE}
\glsxtrnewsymbol[description={Electricity load forecasting}]{EPF}{EPF}
\glsxtrnewsymbol[description={Exponential smoothing models}]{ESM}{ESM}
\glsxtrnewsymbol[description={Extract transform load}]{ETL}{ETL}
\glsxtrnewsymbol[description={Electric vehicles}]{EVs}{EVs}

@@

% T
\glsxtrnewsymbol[description={Hadamard product}]{odot}{\ensuremath{\odot}}
